Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models

### 2015-02-07 processes-69103_review_response/processes-69103.tex
Final changes

Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Khanin:1989aa,Willems:1991aa,Baldwin:1994aa,Leipold:1995aa,Kuharsky:2001aa}.
Mechanistic ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, Hockin \emph{et al.} \citep{Hockin:2002aa}, and later Butenas et al., \citep{Butenas:2004aa}
who developed and then subsequently refined highly mechanistic coagulation models. Recently Papadopoulos and co-workers used a phenomenological  mathematical model for thrombin generation \cite{Atkin:2014}. Using a set of four ordinary differential equations they were able to derive an equation for temporal evolution of thrombin generation. However the model focuses more on thrombin generation rather than dynamics of coagulation. It lacks any detail regarding regulatory activity and is unable to capture the roles of primary inhibitors like ATIII or protein C as the hybrid approach captures.The model also requires that the parameters are adjusted or fine tuned for each data set. This makes the model unfeasible in cases where we need to understand the dynamics of coagulation without complex modeling approaches or when there is incomplete mechanistic knowledge. Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa}, and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
However, these previous studies were largely based upon extensive mechanistic knowledge. 
This is possible because blood, while enormously complex, can be systematically interrogated.
Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which a complete mechanistic understanding is missing. 
We tested this approach by modeling the human coagulation cascade. 
The hybrid model consisted of only five differential equations augmented with several logical rules.
Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

The performance of the proof of principle coagulation model was impressive given its limited size, however, there are several issues that could be further explored.
First, the prediction of initiation time should be investigated. 
We were able to estimate initiation time within a data set, but unable to predict initiation time \textit{across} independent data sets. 
This suggested that we should update the initiation module to distinguish between different triggers, e.g., TF/FVIIa versus rFVIIa alone, 
and to include key biological milestones such as FXa activation (a prerequisite to thrombin formation).
Next, there are several additional biological modules that could be added to the core model presented here. 
First, we could include thrombin-induced platelet activation and the role of activated platelets in amplification.
We captured thrombin generation data in the presence of platelets, however,
the initial shape of the activation curve and the time-scale of activation was not always consistent with the data.
Platelets are activated by thrombin through the cleavage of the extracellular domain of protease-activated receptors (PARs) on the platelet surface.
Once activated, platelets play an important role in amplification, and are key mediators of the positive feedback driving amplification.
Thus, this biology is a potentially important component of an expanded model.
We should also add the intrinsic pathway to the model. 
The intrinsic pathway is triggered by contact activation of the plasma protease factor XI (fXI) by negatively charged surfaces and by thrombin and upstream factors such as activated
plasma protease factor XII (FXIIa) \citep{NAITO1991,Gailani:1991aa}.
Activated platelets may also release polyphosphate which directly activates fXII \cite{Smith:2006zl}. 
Arguably a minor player in acute bleeding, contact activation could also be important in other wound healing contexts. 
Finally, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation and clot dissolution (fibrinolysis).
Clot formation is driven by thrombin activity, while fibrinolysis is driven by plasmin activity, a key enzyme that cleaves fibrin (one of the main materials in a clot).
Similar to coagulation, fibrinolysis is managed by several activating and inhibitory factors which control the balance between clot formation and dissolution.
Tissue plasminogen activator (t-PA) and urokinase activate plasmin, along with contact pathway factors such as fXIa. On the other hand, thrombin activatable fibrinolysis inhibitor (TAFI)
inhibits the degradation of fibrin by plasmin. Also, similar to coagulation, there is considerable fibrinolysis and contact pathway data sets that can be used to train the model.
Lastly, the choice of max/min integration rules or the particular form of the transfer functions could be generalized to include other rule types and functions. 
Theoretically, an integration rule is a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
Alternative integration rules such as the mean might have different properties which could influence model identification or performance. 
For example, a mean integration rule would be differentiable, which allows derivative-based optimization approaches to be used. 
The particular form of the transfer function could also be explored. We choose a Hill-like function because of its 
prominence in the systems and synthetic biology community. 
However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the 
range $f\in[0,1]$. Thus, many types of transfer functions are possible.

>>>>

Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Khanin:1989aa,Willems:1991aa,Baldwin:1994aa,Leipold:1995aa,Kuharsky:2001aa}.
Mechanistic ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, Hockin \emph{et al.} \citep{Hockin:2002aa}, and later Butenas et al., \citep{Butenas:2004aa}
who developed and then subsequently refined highly mechanistic coagulation models. Recently Papadopoulos and co-workers used a phenomenological  mathematical model for thrombin generation \cite{Atkin:2014}. Using a set of four ordinary differential equations they were able to derive an equation for temporal evolution of thrombin generation. However the model focuses more on thrombin generation rather than dynamics of coagulation. Unlike our hybrid approach it does not consider regulatory detail and thus is unable to capture the roles of primary inhibitors like ATIII or protein C. Thus in cases where we need to understand the dynamics of coagulation without complex modeling or when there is incomplete mechanistic knowledge, the hybrid approach is a better candidate. The model also requires that the parameters are adjusted or fine tuned for each data set unlike the hybrid approach where we use a single ensemble across all experiments. However platelet modeling is a significant advantage of this model over the hybrid approach.  Future versions of the hybrid model can involve modeling platelets as done by Papdopoulos et al. in their phenomenological model. Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa}, and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
However, these previous studies were largely based upon extensive mechanistic knowledge. 
This is possible because blood, while enormously complex, can be systematically interrogated.
Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which a complete mechanistic understanding is missing. 
We tested this approach by modeling the human coagulation cascade. 
The hybrid model consisted of only five differential equations augmented with several logical rules.
Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

The performance of the proof of principle coagulation model was impressive given its limited size. The most detailed mechanistic model of coagulation is the one by Luan and co-workers built  with 193 proteins and protein complexes that are interconnected by 301 reactions \citep{2010_luan_varner_MolBioSys}. Akin to the hybrid model this mechanistic used the normal thrombin data and hemophilia data from Allen \emph{et al.} \citep{ALLEN2006}. However unlike the mechanistic model we used this data for validation rather than training. Results from our model are surprisingly comparable to the training simulations of the mechanistic model. The initiation time and amplification of thrombin signal were accurately predicted in the normal case whereas in the hemophilia case we correctly predict the initiation time but slightly over predict the amplification signal. This performance is nearly similar to that of the detailed mechanistic model with 301 reactions. Despite the impressive performance, there are several issues that could be further explored. First, the prediction of initiation time should be investigated. 
We were able to estimate initiation time within a data set, but unable to predict initiation time \textit{across} independent data sets. 
This suggested that we should update the initiation module to distinguish between different triggers, e.g., TF/FVIIa versus rFVIIa alone, 
and to include key biological milestones such as FXa activation (a prerequisite to thrombin formation).
Next, there are several additional biological modules that could be added to the core model presented here. 
First, we could include thrombin-induced platelet activation and the role of activated platelets in amplification.
We captured thrombin generation data in the presence of platelets, however,
the initial shape of the activation curve and the time-scale of activation was not always consistent with the data.
Platelets are activated by thrombin through the cleavage of the extracellular domain of protease-activated receptors (PARs) on the platelet surface.
Once activated, platelets play an important role in amplification, and are key mediators of the positive feedback driving amplification.
Thus, this biology is a potentially important component of an expanded model.
We should also add the intrinsic pathway to the model. 
The intrinsic pathway is triggered by contact activation of the plasma protease factor XI (fXI) by negatively charged surfaces and by thrombin and upstream factors such as activated
plasma protease factor XII (FXIIa) \citep{NAITO1991,Gailani:1991aa}.
Activated platelets may also release polyphosphate which directly activates fXII \cite{Smith:2006zl}. 
Arguably a minor player in acute bleeding, contact activation could also be important in other wound healing contexts. 
Finally, to make the model more clinically relevant, we should include the biochemical processes responsible for clot formation and clot dissolution (fibrinolysis).
Clot formation is driven by thrombin activity, while fibrinolysis is driven by plasmin activity, a key enzyme that cleaves fibrin (one of the main materials in a clot).
Similar to coagulation, fibrinolysis is managed by several activating and inhibitory factors which control the balance between clot formation and dissolution.
Tissue plasminogen activator (t-PA) and urokinase activate plasmin, along with contact pathway factors such as fXIa. On the other hand, thrombin activatable fibrinolysis inhibitor (TAFI)
inhibits the degradation of fibrin by plasmin. Also, similar to coagulation, there is considerable fibrinolysis and contact pathway data sets that can be used to train the model.
Lastly, the choice of max/min integration rules or the particular form of the transfer functions could be generalized to include other rule types and functions. 
Theoretically, an integration rule is a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
Alternative integration rules such as the mean might have different properties which could influence model identification or performance. 
For example, a mean integration rule would be differentiable, which allows derivative-based optimization approaches to be used. 
The particular form of the transfer function could also be explored. We choose a Hill-like function because of its 
prominence in the systems and synthetic biology community. 
However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the 
range $f\in[0,1]$. Thus, many types of transfer functions are possible.
